Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with the R packages \pkg{ACSNMineR} and \pkg{RNaviCell}

### 2016-01-14 jss/article.tex
[skip ci] small corrections

}
\Keywords{biological module, enrichment, depletion, cancer, maps, molecular networks, molecular pathways, \proglang{R}}
\Plainkeywords{biological module, enrichment, depletion, cancer, maps, molecular networks, molecular pathways, R} %% without formatting
%% at least one keyword must be supplied

@@

\section[Introduction]{Introduction}
Biological pathways and networks comprise sets of interactions or functional
relationships, occurring at the molecular level in living cells
\citep{adriaens2008public}.  A large body of knowledge on cellular biochemistry
is organized in publicly available repositories such as the KEGG database
\citep{kanehisa2011kegg}, Reactome \citep{croft2014reactome} and MINT
\citep{zanzoni2002mint}. All these biological databases facilitate a large
spectrum of analyses, improving our understanding of cellular systems. For
instance, it is a very common practice to cross the output of high-throughput
experiments, such as mRNA or protein expression levels, with curated biological
pathways in order to visualize changes, analyze their impact on a network and
formulate new hypotheses about biological processes. Many biologists and
computational biologists establish list of genes of interest (e.g. a list of
genes that are differentially expressed between two conditions, such as normal
vs disease) and then try to see if known biological pathways are enriched with
this list of genes.

We have recently released the Atlas of Cancer Signalling Networks (ACSN), a
web-based database which describes signaling and regulatory molecular processes
that occur in a healthy mammalian cell but that are frequently deregulated
during cancerogenesis \citep{kuperstein2013navicell}.  The ACSN atlas aims to
be a comprehensive description of cancer-related mechanisms retrieved from the
most recent literature. The web interface for ACSN is using the NaviCell
technology, a software framework dedicated to web-based visualization and
navigation for biological pathway maps \citep{kuperstein2013navicell}. This
environment is providing an easy navigation of maps through the use of the
Google Maps JavaScript library, a community interface with a web blog system,
and a comprehensive module for visualization and analysis of high-throughput
data \citep{bonnet2015navicell}.

In this article, we describe two software packages related to ACSN analysis and
data visualization for the popular R statistical enrironment \citep{mainRref,
vance2009data}. The package \pkg{ACSNMineR} is designed for the calculation of
gene enrichment and depletion in ACSN maps, while \pkg{RNaviCell} is dedicated
to data visualization on ACSN maps. Both packages are available on the
Comprehensive R Archive Network
(\url{https://cran.r-project.org/web/packages/ACSNMineR/} and
\url{https://cran.r-project.org/web/packages/RNaviCell/}), and on the GitHub
repository (\url{https://github.com/sysbio-curie/ACSNMineR} and
\url{https://github.com/sysbio-curie/RNaviCell}). For the remainder of this
article, we describe the organization of each package and illustrate their
capacities with several concrete examples demonstrating their capabilities.

@@

Finally, to facilitate visualization of results, \pkg{ACSNMineR} integrates a
representation function based on \pkg{ggplot2} syntax \citep{ggplot2}. It
allows representation of results from \code{enrichment} or
\code{multisample_enrichment} with a limited number of parameters. For
examples, two types of display are available: heat-map tiles or bars. For
multiple samples using a barplot representation the number of rows used can be
provided, all plots will be on the same row otherwise. For the heatmap, the
color of the non-significant modules, and boundaries of the gradient for
significant values can also be tuned.

@@

\begin{table}[h!]
  \centering
  \caption{Overview of results from enrichment analysis without correction.
  Module : name of the module. Mod. size: size of the module. Genes in module:
  genes from input which are found in the module. p-value: uncorrected p-value.
  Test : null hypothesis used, greater is synonym of enrichment.}
  \label{tab:table2}

@@

\begin{figure}[!ht]
  \caption{Gene expression values from a prostate cancer cell line visualized
on the cell cycle map as heat map plots. The figure is a screenshot of the
NaviCell map browser, with the map set at the top (the less detailed) zoom
level. The essential phases of the cell cycle are indicated on the map
(G1/S/G2/M).} 
  \centering
  \includegraphics[width=0.8\textwidth]{figures/heatmap.pdf}
  \label{fig:du145}
\end{figure}

@@

At last, we visualize the mean expression values for ER negative samples for all
genes differentially expressed on the ACSN master map using \code{ACSNMineR}
commands to create heatmaps.

@@

>>>>

}
\Keywords{biological module, biological pathway, enrichment, depletion, cancer, maps, molecular networks, molecular pathways, \proglang{R}}
\Plainkeywords{biological module, biological pathway, enrichment, depletion, cancer, maps, molecular networks, molecular pathways, R} %% without formatting
%% at least one keyword must be supplied

@@

\section[Introduction]{Introduction}
Biological pathways and networks comprise sets of interactions or functional
relationships, occurring at the molecular level in living cells
\citep{adriaens2008public}.  A large body of knowledge on cellular biochemistry
is organized in publicly available repositories such as the KEGG database
\citep{kanehisa2011kegg}, Reactome \citep{croft2014reactome} and MINT
\citep{zanzoni2002mint}. All these biological databases facilitate a large
spectrum of analyses, improving our understanding of cellular systems. For
instance, it is a very common practice to cross the output of high-throughput
experiments, such as mRNA or protein expression levels, with curated biological
pathways in order to visualize the changes, analyze their impact on a network and
formulate new hypotheses about biological processes. Many biologists and
computational biologists establish list of genes of interest (e.g. a list of
genes that are differentially expressed between two conditions, such as normal
vs disease) and then try to see if known biological pathways are enriched with
this list of genes.

We have recently released the Atlas of Cancer Signalling Networks (ACSN), a
web-based database which describes signaling and regulatory molecular processes
that occur in a healthy mammalian cell but that are frequently deregulated
during cancerogenesis \citep{kuperstein2015atlas}.  The ACSN atlas aims to
be a comprehensive description of cancer-related mechanisms retrieved from the
most recent literature. The web interface for ACSN is using the NaviCell
technology, a software framework dedicated to web-based visualization and
navigation for biological pathway maps \citep{kuperstein2013navicell}. This
environment is providing an easy navigation of maps through the use of the
Google Maps JavaScript library, a community interface with a web blog system,
and a comprehensive module for visualization and analysis of high-throughput
data \citep{bonnet2015navicell}.

In this article, we describe two software packages related to ACSN analysis and
data visualization for the popular R statistical enrironment \citep{mainRref,
vance2009data}. The package \pkg{ACSNMineR} is designed for the calculation of
gene enrichment and depletion in ACSN maps (or any user-defined gene set via
the import function), while \pkg{RNaviCell} is dedicated to data visualization
on ACSN maps. Both packages are available on the Comprehensive R Archive
Network (\url{https://cran.r-project.org/web/packages/ACSNMineR/} and
\url{https://cran.r-project.org/web/packages/RNaviCell/}), and on the GitHub
repository (\url{https://github.com/sysbio-curie/ACSNMineR} and
\url{https://github.com/sysbio-curie/RNaviCell}). For the remainder of this
article, we describe the organization of each package and illustrate their
capacities with several concrete examples demonstrating their capabilities.

@@

Finally, to facilitate visualization of results, \pkg{ACSNMineR} integrates a
representation function based on \pkg{ggplot2} syntax \citep{ggplot2}. It
allows representation of results from \code{enrichment} or
\code{multisample_enrichment} with a limited number of parameters. Two types of
display are available: heat-map tiles or bars. For multiple samples using a
barplot representation, the number of rows used can be provided, otherwise all
plots will be on the same row. For the heatmap, the color of the
non-significant modules, and boundaries of the gradient for significant values
can also be tuned.

@@

\begin{table}[h!]
  \centering
  \caption{Overview of the results from enrichment analysis without correction.
  Module : name of the module. Mod. size: size of the module. Genes in module:
  genes from input which are found in the module. p-value: uncorrected p-value.
  Test : null hypothesis used, greater is synonym of enrichment.}
  \label{tab:table2}

@@

\begin{figure}[!ht]
  \caption{Gene expression values from a prostate cancer cell line visualized
on the cell cycle map as heat map plots. The figure is a screenshot of the
NaviCell map browser, with the map set at the top (the less detailed) zoom
level. The essential phases of the cell cycle are indicated on the map
(G1/S/G2/M). Note that on the web browser the map is interactive and the user
can zoom in and out, change the graphical parameters, import additional data
and perform functional analysis.  } 
\centering
  \includegraphics[width=0.8\textwidth]{figures/heatmap.pdf}
  \label{fig:du145}
\end{figure}

@@

At last, we visualize the mean expression values for ER negative samples for all
genes differentially expressed on the ACSN master (global) map using \code{ACSNMineR}
commands to create heatmaps.

@@

\section[Summary and perspectives]{Summary and perspectives}
In this work, we presented the R package \pkg{ACSNMineR}, a novel package for
the calculation of p-values for enrichment or depletion of genes in biological
pathways. The package includes the six large-scale molecular maps and 55
functional modules of the Atlas of Cancer Signaling Networks (ACSN) . Enrichment can be
calculated for those maps and modules with several options to play with, but can
also be calculated for other databases of molecular pathways, that can be
imported from GMT formated files.

We also describe in this work the \pkg{RNaviCell} package, a R package highly
complementary to \pkg{ACSNMineR}. This package is dedicated to create web-based
and interactive data visualization on ACSN maps. Users can use this tools to
represent genes of interest that have been shown to be related to the maps by
calculating enrichment with the  \pkg{ACSNMineR}.  Creating maps with the
graphical user interface of the ACSN website can be a tedious task if the user
has multiple samples or gene lists, and wants to compare their representations
on ACSN maps. The \pkg{RNaviCell} package can be used to automate the process of
creating the graphical representations automatically.

We have shown how the packages \pkg{ACSNMineR} and \pkg{RNaviCell} can be
combined to analyze expression data from breast cancer samples, and also to
analyze the frequency of mutated genes in glioblastoma cancer samples.

Of course, \pkg{ACSNMineR} is not the only R package for enrichment
calculations. For instance, \pkg{GOstats} \citep{falcon2007using} is probably
one of the first packages that was created to calculate enirchment for Gene
Ontology categories.  \pkg{GOstats} can also be used to calculate enrichment for
other biological pathways categories, such as KEGG pathways (by using an
instance of the class \code{KEGGHyperGParams}) or PFAM protein families (using
\code{PFAMHyperGParams}). However, its usage might not be as straightforward as
\pkg{ACSNMineR}, and it does not seems possible to test user-defined biological
pathways. Furthermore, other authors have pointed out that the KEGG database
used by this package has not been updated since 2012. \pkg{clusterProfiler} is a
recent R package released for enrichment analysis of Gene Ontology and KEGG with
either hypergeometric test or Gene Set Enrichment Analysis (GSEA)
\citep{yu2012clusterprofiler}. Via other packages, support for analysis of
Disease Ontology and Reactome Pathways is possible. Interestingly, this package
also offers the possibility to import user-defined gene set, through
tab-delimited pairwise definition files.

In order to improve \pkg{ACSNMineR}, we may in the near future try to improve
the speed of calculation, which might be a problem if a very large number of
samples or experiments have to be analyzed rapidly. For instance, we could use
the \code{foreach} and \code{\%dopar\%} operator to parallelize the most
computationally demanding operations. It could also be useful to implement more
sensitive methods of gene set enrichment measures, such as the Gene Set
Enrichment Analysis (GSEA) method.
